Polynomial Time and Full Dependent Types

### 2023-07-11 paper/paper.tex
[ paper ] some writing, mostly introduction

\documentclass[acmsmall,review,screen]{acmart}
\settopmatter{printfolios=true,printccs=true,printacmref=true}

@@

%%
%% The abstract is a short summary of the work to be presented in the
%% article.
\begin{abstract}
  We combine dependent types with linear type systems that soundly and
  completely capture polynomial time computation. To combine dependent
  and linear types, we use \emph{Quantitative Type Theory}. We explore
  two systems for capturing polynomial time: one system that disallows
  construction of iterable data, and one, based on the LFPL system of
  Martin Hofmann, that controls construction via a payment
  method. Both of these are extend to full dependent types, allowing
  for arbitrary computation in types alongside guaranteed polynomial
  time computation in terms.

  The resulting system smoothly combines reasoning with dependent
  types and restricted computational complexity. \bob{Elaborate after
    Section 3 is written.}We prove the soundness of the systems using
  a realisability technique due to Hofmann and Dal Lago.
\end{abstract}

@@

{\bf New story:}
\begin{enumerate}
\item Introduction: The goal of this paper is to unify implicit
  computational complexity with full dependent types. ICC enables us
  to ensure that \emph{programs} are within a certain complexity
  class, while dependent types allow us to state and prove that these
  programs solve the problems that we intend them to, and to construct
  reductions between problems. We can marry complexity-controlled
  implementations with problems, paving the way for a \emph{synthetic
    computational complexity theory}.
\item Section 1: Linear types and Implicit Polynomial Time.
  \begin{enumerate}
  \item Total typed functional programming, as we see in type theory,
    is restricted to the use of higher order functions and iteration
    over inductive data types.
  \item Polynomial time is the smallest class closed under nested
    iteration over the input -- i.e. we consider iteration over the
    input as feasible, and compositions thereof. The key thing is to
    avoid iteration over data constructed from the input.
  \item Key is to control iteration. So we split inductive datatypes
    into iterable and non-iterable versions. We can
  \item One way to capture polytime is to ensure that all iteration is
    only done over the input. So we can duplicate iterable things, but
    we can't construct new iterable things (at least not from an
    iteration). We also can't duplicate higher order things, because
    that effectively acts as a way to duplicate iterations (half
    instantiated iterators). Compare to Jones' CONS-FREE programming.
  \item Another approach is to allow ``borrowing'' of iterability by
    means of diamonds. Iterability becomes a first-class object, so we
    can use it to construct new data structures from the input. This
    allows the construction of new iterable data structures. LFPL.
  \end{enumerate}
\item Section 2: Extending to Dependent Types
  \begin{enumerate}
  \item Combining linear types and dependent types is subtle, because
    there is a fundamental incompatibility in the syntax of the two
    systems.
  \item There have been several approaches, though they are all
    somewhat similar (list them all here). We will use QTT, because it
    offers a firm semantic foundation as well as a relatively
    straightforward syntax.
  \item Recap the syntax of QTT, with extensional equality. Key point
    is that there is a normal dependently typed language with a linear
    sub-language. We can have special constructs in the linear
    sub-language that ``erase'' to normal type theory -- products are
    the prime example.
  \item Non iterable datatypes -- have normal eliminators in the DTT
    fragment, but
  \end{enumerate}
\item Section 3: Programming and Proving with Poly-time
  \begin{enumerate}
  \item Internalised completeness proofs.
  \item Construct the type of problems that are solvable in polynomial
    time.
  \item Construct reductions between problems.
  \item NP. Doesn't make sense to talk about this without dependent
    types, because we can't say when we've solved the problem being
    stated.
  \item PP (Probabilistic), using a monad. Again, need to be able to
    state connection with the problem to know that we are solving it.
  \end{enumerate}
\item Section 4: Soundness via Realisability
  \begin{enumerate}
  \item The underlying CBV language and its
  \end{enumerate}
\item Section 5: Extensions
  \begin{enumerate}
  \item Explicit Complexity via Diamond Counting
  \item Poly-time univalence
  \item Towards P- and NP-completeness proofs. Issue here is the
    extensionality. Towards a synthetic (poly-time based)
    computational complexity theory.
  \end{enumerate}
\item Section 6: Related Work
  \begin{enumerate}
  \item ICC stuff
    \begin{enumerate}
    \item LFPL
    \item CONS-free
    \item Soft Affine Logic; which has the same underlying idea as
      CONS-free (is this a novel observation?)
    \item Other systems based on information flow.
    \end{enumerate}
  \item Explicit complexity: lambda-amor, and things by Deepak Garg
    and Marco Gaboardi.
  \item Recent CMU stuff on calf
  \end{enumerate}

\end{enumerate}

{\bf Old story:}
\begin{enumerate}
\item There are systems for implicit computational complexity based on
  linear logic (e.g. LFPL, SAL, LLL, BLL) that seek to capture
  complexity classes (most often polynomial time) implicitly
  (i.e. without directly talking about time bounds). This work has
  also led to followup work on explicitly resourced calculi such as
  RAML, which seek to use static analysis to determine complexity
  bounds on programs.
\item In dependent type theory, we can reason about the extensional
  behaviour of programs (within the limits of intensional type
  theory), but not about intenstional properties. The only way to
  formalise something like ``for all polynomial time functions
  $\mathbb{N} \to \mathbb{N}$'' is to develop a deep emebdding of
  polynomial time functions and to work with that, which is tedious.
\item The goal of this paper is to unify dependent types for reasoning
  about programs with implicit computational complexity type
  systems. In the first instance, we do not seek to use dependent
  types to state precise bounds on programs' time consumption, but we
  intend to use dependent types to reason about the extensional
  behaviour of programs that we know to be polynomial time. At the end
  of the paper, we show how to adapt our framework to explicit
  resource accounting, in the style of RAML.
\item The plan of the paper is as follows:
  \begin{enumerate}
  \item We first introduction the ideas behind the use of linear logic
    for capturing polynomial time, by studying two systems Lafont's
    SAL, and Hofmann's LFPL. In Hofmann and Dal Lago's presentation,
    these are two variants of Second order Multiplicative Linear
    Logic, with restricted $\oc$ modalities for duplication. We also
    recall the realisability semantics of Dal Lago and Hofmann (FIXME:
    move this later?)
  \item The most modern presentations of these systems use second
    order quantification to represent data types. This is incompatible
    with the form of dependent types we use here, so we seek a direct
    representation of natural numbers that captures
  \item We then recall QTT, a extension of Martin-Löf type theory with
    linear types, which can accomodate intensional restrictions on
    computability as well as full reasoning about programs. Since
    allowing an unrestricted $\oc$ modality would violate our
    polynomial time bounds, we instead choose the natural number
    semiring for resource accounting.
  \item We then transplant the simply typed rules into QTT, showing
    how to carefully distinguish between data that is available to the
    program, and data that is only available for type formation.
  \item We finish by showing that we can also do explicit resource
    tracking.
  \end{enumerate}
\end{enumerate}

@@

For this section, the linear $\lambda$-calculus we will use will have
linear functions and $\otimes$-products:
\begin{mathpar}
  \inferrule*
  { }
  {x : A \vdash x : A}

@@

  \inferrule*
  {\Gamma_1 \vdash M : A \otimes B \\ \Gamma_2, x : A, y : B \vdash N : C}
  {\Gamma_1, \Gamma_2 \vdash \Let\,(x,y)=M\,\In\,N : C}
\end{mathpar}
These rules are standard, so we do not describe them further except to
note how linear typing uses presence or absences in a context to
control resource usage -- if a variable is in the context it must be
used exactly once. The fact that this discipline interferes with
dependent types is one of the reasons we turn to QTT when we wish to
add dependent types in \autoref{sec:qtt}.

@@

  \inferrule*
  {\Gamma \vdash M \stackrel0: (x \stackrel\pi: S) \otimes T}
  {\Gamma \vdash \mathrm{snd}(M) \stackrel0: T[\mathrm{fst}(M)/x]}
\end{mathpar}
Hence, $\Sigma$-types act as normal in the $\sigma = 0$ fragment, and
we can use them as for normal type-theoretic constructions.  In the
$\sigma = 1$ fragment, we must take into account the resource content
of objects and use a pattern matching construct; the dependently typed
analogue of the $\otimes$-eliminator in \autoref{sec:lineartype}:
\begin{mathpar}
  \inferrule*
  {0\Gamma, z \stackrel0: (x \stackrel\pi: A) \otimes B \vdash C \\
    \Gamma_1 \vdash M \stackrel\sigma: (x \stackrel\pi: A) \otimes B \\
    \Gamma_2, x \stackrel{\sigma\pi}: A, y \stackrel{\sigma}: B \vdash N \stackrel\sigma: C[(x,y)/z]\\
    0\Gamma_1 = 0\Gamma_2}
  {\Gamma_1 + \Gamma_2 \vdash \mathrm{let}~(x,y) = M~\mathrm{in}~N \stackrel\sigma: C[M/z]}
\end{mathpar}

@@

\subsubsection{Universe} QTT can also be equipped in universe types,
just as in standard type theory. We refer to the rules presented in
\cite{atkey18qtt}.

@@

\subsection{Polytime Problems}

@@

\paragraph{Alternative definition} Every resource monoid induces a
partial ordering on its carrier set by $\alpha \leq \beta$ iff
$0 \leq M(\alpha, \beta)$. Taking this idea further, we can
reformulate a resource monoid as a symmetric monoidal category
enriched in the symmetric monoidal category $\natinf$, where the
monoid structure is addition. Isomorphism of objects $\alpha, \beta$
in such a category amounts to having $M(\alpha, \beta) = 0$. The
conditions in the definition above amount to the usual identity and
composition laws for enriched categories.

@@

\paragraph{Agda Formalisation} The natural number resource monoid is
defined in the Agda module \AgdaModule{Algebra.ResourceMonoid.Nat} and the
 polynomial monoids are both defined in
\AgdaModule{Algebra.ResourceMonoid.Polynomial}. The definition is
parameterised by the ``size monoid'' operation (either $\sqcup$ or
$+$) used to compose sizes.

@@

\begin{enumerate}
\item RAML, Hofmann-Jost, etc.
\item lambda-amor
\item
\end{enumerate}

@@

@@

@@

>>>>

\documentclass[acmsmall,review,screen,anonymous]{acmart}
\settopmatter{printfolios=true,printccs=true,printacmref=true}

@@

%%
%% The abstract is a short summary of the work to be presented in the
%% article.
\begin{abstract}
  We combine dependent types with linear type systems that soundly and
  completely capture polynomial time computation. To combine dependent
  and linear types, we use \emph{Quantitative Type Theory}. We explore
  two systems for capturing polynomial time: one system that disallows
  construction of iterable data, and one, based on the LFPL system of
  Martin Hofmann, that controls construction via a payment
  method. Both of these are extend to full dependent types, allowing
  for arbitrary computation in types alongside guaranteed polynomial
  time computation in terms. We prove the soundness of the systems
  using a realisability technique due to Hofmann and Dal Lago.

  Our long-term goal is to combine the extensional reasoning of type
  theory with intensional reasoning about the resources intrinsically
  consumed by programs. This paper is a step along this path, which we
  hope will lead both to practical systems for reasoning about
  programs' resource usage, and to theoretical use as a form of
  \emph{synthetic computational complexity theory}.
\end{abstract}

@@

Type theory is often claimed to be ideal for Computer Science,
combining programming and proof in one unifying system, so a happy
programmer can verify while they program, and program while they
verify. From a broader Computer Science view, however, type theory
lacks the ability to talk about the very thing that makes Computer
Science interesting -- the fact that computation is everywhere bounded
by the resources in time and space that we can afford to give it.

Typical Type Theory only speaks of the public face that programs
present to the world -- if you input things like this, you get things
that look like that -- but cannot bring itself to mention the true
cost of programs' execution. One can encode costs by embedding another
programming language in type theory \cite{pottier}, or one can
synthesise costs by treating resource counting as a computational
effect \cite{daniellsson,calf}, but neither of these capture the
intrinsic costs of the programs we write in type theory. These
techniques deliver only \emph{conspicuous consumption}, not speaking
of the real resources consumed.

In this paper, we propose a method for extending dependent type theory
with a means for constraining the intrinsic computational complexity
of programs written in the theory. We concentrate on linear type
systems that soundly and completely capture polynomial time
computation, the commonly used standard for feasible resource usage,
and extend these systems to dependent types. The additional
expressivity of dependent types allows us use these characterisations
of polytime to further functionally characterise

We use techniques from \emph{Implicit Computational Complexity}
theory, which seeks to provide intrinsic characterisations of
complexity classes in terms of logical systems or programming
languages. We review the techniques that we use in
\autoref{sec:linear-intro}. To adapt these systems to dependent types,
we use \emph{Quantitative Type Theory} (QTT)
\cite{atkey18qtt,mcbride16}, a combination of linear and dependent
types. We review QTT in \autoref{sec:qtt}.

Our long-term goal is to combine the extensional reasoning of type
theory with intensional reasoning about the resources intrinsically
consumed by programs. This paper is a first step along this path,
which we hope will lead both to practical systems for reasoning about
programs' resource usage, and to theoretical use as a form of
\emph{synthetic computational complexity theory}. We discuss these
possibilities further in \autoref{sec:conclusion}.

% The goal of this paper is to unify implicit computational complexity
% with full dependent types. ICC enables us to ensure that
% \emph{programs} are within a certain complexity class, while dependent
% types allow us to state and prove that these programs solve the
% problems that we intend them to, and to construct reductions between
% problems. We can marry complexity-controlled implementations with
% problems, paving the way for a \emph{synthetic computational
%   complexity theory}.

% \item Section 5: Extensions
%   \begin{enumerate}
%   \item Explicit Complexity via Diamond Counting
%   \item Poly-time univalence
%   \item Towards P- and NP-completeness proofs. Issue here is the
%     extensionality. Towards a synthetic (poly-time based)
%     computational complexity theory.
%   \end{enumerate}
% \item Section 6: Related Work
%   \begin{enumerate}
%   \item ICC stuff
%     \begin{enumerate}
%     \item LFPL
%     \item CONS-free
%     \item Soft Affine Logic; which has the same underlying idea as
%       CONS-free (is this a novel observation?)
%     \item Other systems based on information flow.
%     \end{enumerate}
%   \item Explicit complexity: lambda-amor, and things by Deepak Garg
%     and Marco Gaboardi.
%   \item Recent CMU stuff on calf
%   \end{enumerate}

\subsection{Contributions and Content}

This paper makes the following contributions to the theory and use of
linear dependent type theory and implicit computational complexity:
\begin{enumerate}
\item We formulate two systems that combine linear type theory for
  polytime computation with full dependent types, using Quantitative
  Type Theory. The system is presented in \autoref{sec:qtt}. The
  linear typing discipline required for enforcing polytime is provided
  by QTT, but we also need to carefully add constructs for
  non-iterable datatypes (\autoref{sec:noniter-qtt}) and the two kinds
  of natural number iterator that we consider
  (\autoref{sec:cons-free-qtt} and \autoref{sec:lfpl-qtt}). Porting
  the natural number iterators from the simply typed to the
  dependently typed setting requires careful annotation of the rules
  to ensure that the correct information is available for type
  checking, while also not allowing too much information to be made
  available at runtime that would violate the polytime soundness
  property. A further contribution of this paper is the addition of
  reflection types to QTT, \autoref{sec:qtt-reflection}, which allow
  statements about polytime realisability to be reflected into
  types. We explain in \autoref{Sec:realisability-future} how this
  will have applications in QTT beyond the polytime systems presented
  here.
\item We demonstrate the utility of the combination of polytime and
  dependent types in \autoref{sec:programming-polytime}. Just as in
  the simply typed world, we have an expressive language for writing
  polytime programs. With the additional power of dependent types, we
  can also prove properties of these programs. A simple example is
  proving that a sorting program actually sorts. Using QTT reflection,
  we can go further and represent the class of polytime problems, with
  polytime reductions between them, as dependent pairs
  (\autoref{sec:polytime-problems}). Our final examples use dependent
  types to give monadic presentations of the complexity classes of
  Non-deterministic Polynomial (NP) time and Probabilistic Polynomial
  (PP) time. Since these classes rely on specific correctness
  criteria, it is not possible to capture them in a simply typed
  system for polytime.
\item We prove the polytime soundness of our systems via a
  realisability argument in \autoref{sec:soundness}. Our construction
  is an extension of the amortised complexity realisability
  constructions of \cite{hofmann-dallago}. We extend their work to our
  dependently typed setting, and also give a realisability
  interpretation of datatypes directly, instead of via second-order
  impredicative encodings. The technical content of this section has
  been formalised in the Agda proof assistant \cite{agda}, and the
  development is available in the supplementary material.
\end{enumerate}

Before we get to the contributions above, we present, in
\autoref{sec:linear-intro}, two linear simply typed systems for
polytime, adaptations of systems already present in the
literature. Our paper concludes with a discussion of further related
work and the outlook for future work in \autoref{sec:conclusion}.

@@

For this section, the affine linear $\lambda$-calculus we will use
will have linear functions and $\otimes$-products:
\begin{mathpar}
  \inferrule*
  { }
  {\Gamma, x : A, \Gamma' \vdash x : A}

@@

  \inferrule*
  {\Gamma_1 \vdash M : A \otimes B \\ \Gamma_2, x : A, y : B \vdash N : C}
  {\Gamma_1, \Gamma_2 \vdash \Let\,(x,y)=M\,\In\,N : C}
\end{mathpar}
These rules are standard, so we do not describe them further except to
note how affine linear typing uses presence or absences in a context
to control resource usage. If a variable is in the context it must be
used at most once (variables that are not used are absorbed by the
additional context in the variable rule). The fact that this
discipline interferes with dependent types is one of the reasons we
turn to QTT when we wish to add dependent types in \autoref{sec:qtt}.

@@

  \inferrule*
  {\Gamma \vdash M \stackrel0: (x \stackrel\pi: S) \otimes T}
  {\Gamma \vdash \mathrm{snd}(M) \stackrel0: T[\mathrm{fst}(M)/x]}
\end{mathpar}
Hence, $\Sigma$-types act as normal in the $\sigma = 0$ fragment, and
we can use them as for normal type-theoretic constructions.  In the
$\sigma = 1$ fragment, we must take into account the resource content
of objects and use a pattern matching construct; the dependently typed
analogue of the $\otimes$-eliminator in \autoref{sec:linear-intro}:
\begin{mathpar}
  \inferrule*
  {0\Gamma, z \stackrel0: (x \stackrel\pi: A) \otimes B \vdash C \\
    \Gamma_1 \vdash M \stackrel\sigma: (x \stackrel\pi: A) \otimes B \\
    \Gamma_2, x \stackrel{\sigma\pi}: A, y \stackrel{\sigma}: B \vdash N \stackrel\sigma: C[(x,y)/z]\\
    0\Gamma_1 = 0\Gamma_2}
  {\Gamma_1 + \Gamma_2 \vdash \mathrm{let}~(x,y) = M~\mathrm{in}~N \stackrel\sigma: C[M/z]}
\end{mathpar}

@@

\subsubsection{Universe} QTT can also be equipped in universe types
$\mathsf{U}$, just as in standard type theory. We refer to the rules
presented in \cite{atkey18qtt}.

@@

\subsection{Polytime Problems}
\label{sec:polytime-problems}

@@

\paragraph{Alternative definition} Every resource monoid induces a
partial ordering on its carrier set by $\alpha \leq \beta$ iff
$0 \leq M(\alpha, \beta)$. Taking this idea further, we can
reformulate a resource monoid as a symmetric monoidal category
enriched in the symmetric monoidal category $\natinf$, where the
monoid structure is addition. The conditions in the definition above
amount to the usual identity and composition laws for enriched
categories.

@@

\paragraph{Agda Formalisation} The $\mathbb{N}$ resource monoid is
defined in \AgdaModule{Algebra.ResourceMonoid.Nat} and the polynomial
monoids are both defined in
\AgdaModule{Algebra.ResourceMonoid.Polynomial}. The definition is
parameterised by the ``size monoid'' operation (either $\sqcup$ or
$+$) used to compose sizes.

@@

\begin{enumerate}
\item RAML, Hofmann-Jost, Raja, etc.
\item lambda-amor,
\item calf from CMU
\end{enumerate}

@@

What we didn't do: polytime typechecking guarantees.

@@

\subsection{Towards Applications}

Cryptography?

@@

Synthetic computability theory Bauer, Forster
